One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2021-12-13 paper.tex
address Rick's comments

The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}. Here we present an analysis of wildlife survey data collected on a critically endangered \hawaii{}an forest bird that aims to meet this objective.  The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.  This information is critical when decisions about conservation strategies are made.  However, estimating the abundance and spatial distribution of wild populations of animals presents many statistical challenges.

@@

Third, a key aim of the analysis is to predict animal density at unsurveyed locations.  If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed location.  However, in many ecological contexts there can be good reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g. the `sociality' of the species).  For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient for which no straightforward explanatory cause has been found \cite{camp_dsm_2020}.  From a statistical perspective, this heterogeneity that cannot be explained by available covariates suggests the inclusion of spatially structured random effects in the model.  Such effects require a careful consideration of the model structure and inference methods to ensure that model fitting remains computationally feasible.

Lastly, the above challenges mean the resulting statistical model is necessarily complex. It is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.  The \akepa{} data is collected with the clear objective of monitoring the population and informing conservation management decisions.  These decisions will be taken by stakeholders with a range of statistical expertise.  This final step of communicating results should therefore be considered with as much care as the rest of the analysis.  This requires the input of statisticians to ensure uncertainty is appropriately captured and communicated in the key outputs of the analysis upon which decisions will be based.  In particular, in the context of species distribution modelling, we note the challenge of communicating uncertainty in maps of predicted animal density.

@@

% Update to section refs
The rest of the paper proceeds as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the GMRF random effect; and Sections \ref{sec-results} and \ref{sec-comms} present the results of the analysis and discuss model evaluation and communication.

@@

The survey data in the open forest stratum is available for download \citep{camp_datarelease_2002} and the full time series data was analysed in a spatio-temporal context in \citep{camp_dsm_2020}.  For the purposes of our analysis we select a single survey year from the \akepa{} time series (2002) that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability. In 2002, there were a total of 289 point transects sampled within the 4,603 ha study area of Hakalau. (Fig. \ref{fig:2002studyareapointspt}) A total of 276 \akepa{} were detected on 121 point transects. The number of detections within each point transect ranged from zero to six. We selected data from a single year to demonstrate our approach in a simplified setting that does not require a temporal model component.  The choice of year is somewhat arbitrary and the model has also been tested analysing data from other years with similar results to those presented here.  However, our approach is readily extendible to a multi-year analysis, similar to that in \cite{camp_dsm_2020}, which we discuss in more detail in Section \ref{sec-discussion}.

@@

Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and a parametric detection function to estimate the detectability of animals as a function of distance from transect or observer \citep{buckland_advanced_2004, buckland_distance_2015}.  Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.  The probability of detection is modelled using a parametric equation and, given estimates of detectability, a randomized sampling design allows for the construction of Horvitz-Thompson-like estimators of animal density \citep{ buckland_advanced_2004, horvitz_generalization_1952}.  Under this approach, animal density is uniform within each strata of the survey.

More recently, interest has focused on fully model-based approaches that model a spatially varying animal density and allow for the use of non-random survey designs.  These methods allow animal density to be associated with spatially-indexed covariates and non-parametric random effects.  Therefore, animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.  As such, varying density estimates are not restricted by the stratification design required for Horvitz-Thompson-like estimators.  This additional flexibility to make predictions for smaller, user-specified geographic units, and extrapolate relationships to unsurveyed regions has made model-based distance sampling a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.

@@

A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the spatial model.  Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties with the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_VariancePropagationDensity_2021, williams_chilean_2011}. Instead, \cite{bravington_VariancePropagationDensity_2021} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of the detection function at the first-stage maximum-likelihood estimate.

Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.  A popular one-stage maximum-likelihood method, introduced by \cite{royle_ModelingAbundanceEffects_2004} and implemented in the \texttt{R} package \texttt{unmarked} \citep{fiske_UnmarkedPackageFitting_2011}, is based on reformulating the model as a multinomial likelihood.  This depends choosing discrete distance classes and binning the data into counts defined on discrete spatial units.  Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, a Poisson likelihood that can be optimised to achieve one-stage inference.  This approach rests heavily on the discretisation of both the distance data and the spatial location data.  Below we present a point process perspective that does not require this discretisation step in either the distance or spatial location data.  The \texttt{unmarked} package also does not implement spatially structured random effects, something that we use here and is also popular in the two-stage approach implemented using \texttt{mgcv}.

@@

A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, usually a key feature of detection functions.  The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful. Representing the observation model as an SPDE also differs substantially from standard distance sampling models that use a parametric model for the detection function.  This may have hindered a wider adoption of using INLA for distance sampling analyses, \cite{yuan_point_2017} remains the only example we are aware of.  Furthermore, parametric forms of detection functions are readily extendible by including covariates and random effects to explain variation in detectability.  It is not clear how to do this the SPDE formulation.  These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA and the benefits of a one-stage modelling process.

Here we present an approach that is closely related to \citet{yuan_point_2017}.  We also fit the model using INLA and take advantage of the benefits the increased computational efficiency compared to MCMC.  However, we avoid the issues with the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal detection function, as will be more familiar to users of classical distance sampling methods. However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \texttt{R-INLA} implementation of INLA \citep{rue_approximate_2009}.  We address this by using a method of iterated model fits based on a first-order Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019}.

We fit the model to point transect \akepa{} data, which differ from line transect methods in that there is a need to account for the increasing area surveyed as distance increases from the transect.  Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process viewpoint is not new and has been taken numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.

@@

In this analysis we assume a half-normal detection function with no covariates on detection parameters to demonstrate the approach in a simplified setting.  The detection function parameters can only be estimated if an assumption is made about $\lambda(\bs)$ otherwise $g(\bs)$ and the intensity are confounded.  The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(\bs)$. The usual approach for two stage models is to assume $\lambda(\bs)$ is constant within each spatial unit, however we are able to relax this assumption slightly to allow $\lambda(\bs)$ to be a linear function within in each transect.  Given such an assumption, any observed deviations from uniformity can be attributed to detectability and not to variation in the intensity.

@@

\sloppy The likelihood \eqref{lgcp-likelihood} contains an integral that cannot be solved analytically.  A common approach for fitting log-Gaussian Cox processes in practice is to approximate the likelihood by replacing the integral with a weighted sum.  In this section we describe this approximate likelihood and show how it can be written as a modified Poisson likelihood.  We follow closely with the approach given in \cite{simpson_going_2016} for a fully observed log-Gaussian Cox process but adapted for the point transect distance sampling case.

To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.  In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes any location within the transect.  It should be clear from context which is intended by $\bs_k$.   The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$.

@@

The approach taken here is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although without the assumption that the locations of observed points form a part of the quadrature scheme.

@@

We denote by $\bar{p}_{\bm{v}}$ the posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.  New linearisation points are found by considering the functional
\begin{equation}
	f(\bar{p}_{\bm{v}}) = \left\{ \argmax_{\bm{u}_i} \: \bar{p}_{\bm{v}}(u_i | \bm{Y}), \quad i = 1, \ldots, n_{\bm{u}} \right\}
\end{equation}
where $n_{\bm{u}}$ is the length of $\bm{u}$.  Each marginal posterior $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ and their modes are estimated during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost for each fit of the linearised model.
We then seek a fixed point of this functional so that $\bm{u}_0 = f(\bar{p}_{\bm{u}_0})$.  This fixed point is identified via an iterative scheme.  Given a current choice of linearisation point $\bm{v}$, estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.  We iterate this procedure until a fixed point is identified within a chosen tolerance.

@@

This section describes the spatial model for the intensity process.  
As mentioned above, the \akepa{} population exhibits spatially heterogeneity for which there is no known explanatory covariate.  Therefore we use a spatially structured random effect to explain this variation.

Using spatially-structured random effects can be computationally expensive.  Often this is due to parameterisations and approximations that lead to dense prior precision matrices which have an  $\mathcal{O}(k^3)$ cost to factorise for an effect with $k$ basis parameters.  A key development in the development of computationally efficient methods is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).  These result in a sparse precision matrix which can be factorised with a cost of $\mathcal{O}(k^{3/2})$.  We use a GMRF representation based on a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.  This representation is constructed by applying finite element methods to solve the SPDE numerically.

@@

In order to specify priors on the Mat\'ern covariance below we use a reparameterisation of $\kappa$ and $\tau$ to a range and variance parameter.  When $\nu = 1$ and the domain is two dimensional then the reparameterisation is $\rho = \sqrt{8} / \kappa$ for the range and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ for the marginal variance (\cite{blangiardo_spatial_2013}, p196).  This parameterisation is used to set penalised complexity priors \citep{simpson_penalising_2017} on $\rho$ and $\sigma^2$ which allow the GRMF effect to shrink towards zero.  For the Mat\'ern field, the base model is the limiting case with $\sigma^2 \rightarrow 0$ and $\rho \rightarrow \infty$, which defines a random field that is almost surely zero everywhere.  The purpose of penalised complexity priors is to place sufficient prior probability on a simple model to allow the random effect to be essentially `turned off' if the model component is not supported by the data.  To specify penalised complexity priors on the SPDE effect we set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.  The value of 130 for the range parameter was selected based on the minimum distance between sampling locations.

@@

The predicted mean of the posterior intensity field is shown in \autoref{fig:intensity-mean-cv}A along with a map of the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid with each prediction cell area approximately 1.7 hectares.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$.}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data \citep{camp_dsm_2020}.  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity.  There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north.  In the north, estimated intensities are lower, which also contributes to larger CV values and makes the CV map hard to interpret.  The standard deviation map shows stronger overall posterior variability in the south where intensity is larger, most likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.

We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location (Figure \ref{fig:intensity-quantiles}).  As we discuss below, these quantiles are independent summary statistics of the posterior intensity field at each prediction location.  
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_quantiles.png}
		\caption{The predicted posterior intensity quantiles:  A) 0.025 quantile B) 0.975 quantile.  Both plots plots use the same colour scale}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}

The posterior detection function (\autoref{fig:det-matern}) shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres and also broadly agrees with a two-stage analysis.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/detfn_and_matern.png}
		\caption{A) Posterior half-normal detection function. B) Posterior Mat\'ern correlation function.  The black line is the mean posterior estimate and the grey shaded area is 95\% credible interval in both plots}
		\label{fig:det-matern}
	\end{center}
\end{figure}
The posterior Mat\'ern correlation function is shown in \autoref{fig:det-matern}; the grey regions show the 95\% credible interval.  We note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).  This is because this is the correlation range on the log-intensity scale.  On the intensity scale the empirical range of correlation is shorter.

@@

To plot the posterior abundance estimate for the \akepa{} in Hakalau reserve we use a Monte Carlo method based on sampling realisations of the posterior intensity field.  Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$ its corresponding random variable within our Bayesian framework.  Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e. the expected abundance.  To generate a full posterior estimate for abundance we approximate the posterior $\pi(N | \bm{Y})$ by taking $m$ Monte Carlo samples of the posterior intensity field, $\lambda^{(1)}, \ldots, \lambda^{(m)}$ and approximate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)}, \bm{Y})$. Each $\pi(N | \lambda = \lambda^{(i)}, \bm{Y})$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda^{(i)}(\bs)\mathrm{d}\bs$. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/N_posterior.png}
		\caption{Posterior abundance.  Grey area marks two Monte-carlo standard errors above and below the mean}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}

@@

The abundance estimates cannot be directly compared to the results in \cite{camp_dsm_2020} since their two-stage model also includes a temporal correlation and we include an area of Hakalau that they had to exclude.  Converting the above abundance estimates to density per hectare for the whole study region allows for some comparison with Figure 2 in \cite{camp_dsm_2020} which suggests that \akepa{} density was roughly between 0.5 and 1.5 birds per hectare in 2002 (considering both the design-based estimate and the spatio-temporal GAM estimate).  In our analysis we find a credible region of roughly 0.9 to 1.7 birds per hectare, which is slightly different but overlaps significantly with the results in \cite{camp_dsm_2020}.

>>>>

The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}. Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.  The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.  This information is critical when decisions about conservation strategies are made.  However, estimating the abundance and spatial distribution of wild populations of animals presents many statistical challenges.

@@

Third, a key aim of the analysis is to predict animal density at unsurveyed locations.  If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed location.  However, in many ecological contexts there can be good reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g. the `sociality' of the species).  For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient for which no straightforward explanatory cause has been found \citep{camp_dsm_2020}.  From a statistical perspective, this heterogeneity that cannot be explained by available covariates suggests the inclusion of spatially structured random effects in the model.  Such effects require a careful consideration of the model structure and inference methods to ensure that model fitting remains computationally feasible.

Lastly, the above challenges mean the resulting statistical model is necessarily complex. It is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.  The \akepa{} data are collected with the clear objective of monitoring the population and informing conservation management decisions.  These decisions will be taken by stakeholders with a range of statistical expertise.  This final step of communicating results should therefore be considered with as much care as the rest of the analysis.  This requires the input of statisticians to ensure uncertainty is appropriately captured and communicated in the key outputs of the analysis upon which decisions will be based.  In particular, in the context of species distribution modelling, we note the challenge of communicating uncertainty in maps of predicted animal density.

@@

The rest of the paper proceeds as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the GMRF random effect; and Sections \ref{sec-results} presents results of the analysis and discusses model evaluation and effective communication of uncertainty.

@@

The survey data used in this analysis is available for download \citep{camp_datarelease_2002} and the full time series data was analysed in a spatio-temporal context in \citep{camp_dsm_2020}.  For the purposes of our analysis we select a single survey year from the \akepa{} time series (2002) that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability. In 2002, there were a total of 289 point transects sampled within the 4,603 ha study area of Hakalau (Fig. \ref{fig:2002studyareapointspt}).  A total of 276 \akepa{} were detected on 121 point transects. The number of detections within each point transect ranged from zero to six. We selected data from a single year to demonstrate our approach in a simplified setting that does not require a temporal model component.  The choice of year is somewhat arbitrary and the model has also been tested analysing data from other years with similar results to those presented here.  However, our approach is readily extendible to a multi-year analysis, similar to that in \cite{camp_dsm_2020}, which we discuss in more detail in Section \ref{sec-discussion}.

@@

Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and a parametric detection function to estimate the detectability of animals as a function of distance from transect or observer \citep{buckland_distance_2015}.  Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.  The probability of detection is modelled using a parametric equation and, given estimates of detectability, a randomized sampling design allows for the construction of Horvitz-Thompson-like estimators of animal density \citep{ buckland_advanced_2004, horvitz_generalization_1952}.  Under this approach, animal density is uniform within each strata of the survey.

More recently, interest has focused on fully model-based approaches that model a spatially varying animal density and allow for the use of non-random survey designs.  These methods allow animal density to be associated with spatially-indexed covariates and non-parametric random effects.  Therefore, animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{buckland_model-based_2016, miller_spatial_2013, johnson_model-based_2010}.  As such, varying density estimates are not restricted by the stratification design required for Horvitz-Thompson-like estimators.  This additional flexibility to make predictions for smaller, user-specified geographic units, and extrapolate relationships to unsurveyed regions has made model-based distance sampling a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.

@@

A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the spatial model.  Early attempts to address this focused on bootstrapping \citep{hedley_spatial_2004, lahiri_resampling_2003} but more recent work has pointed to potential difficulties with the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_VariancePropagationDensity_2021, williams_chilean_2011}. Instead, \cite{bravington_VariancePropagationDensity_2021} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of the detection function at the first-stage maximum-likelihood estimate.

Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.  A popular one-stage maximum-likelihood method, introduced by \cite{royle_ModelingAbundanceEffects_2004} and implemented in the \texttt{R} package \texttt{unmarked} \citep{fiske_UnmarkedPackageFitting_2011}, is based on reformulating the model as a multinomial likelihood.  This depends on choosing discrete distance classes and binning the data into counts defined on discrete spatial units.  Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, a Poisson likelihood that can be optimised to achieve one-stage inference.  This approach rests heavily on the discretisation of both the distance data and the spatial location data.  Below we present a point process perspective that does not require this discretisation step in either the distance or spatial location data.  The \texttt{unmarked} package also does not implement spatially structured random effects, something that we use here and is also popular in the two-stage approach implemented using \texttt{mgcv}.

@@

A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, usually a key feature of detection functions.  The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful. Representing the observation model as an SPDE also differs substantially from standard distance sampling models that use a parametric model for the detection function.  Furthermore, parametric forms of detection functions are readily extendible by including covariates and random effects to explain variation in detectability.  It is not clear how to do this using the SPDE implementation in INLA.  These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA and the benefits of a one-stage modelling process.

Here we present an approach that is related to \citet{yuan_point_2017} but with some key differences. We also fit the model using INLA and take advantage of the benefits the increased computational efficiency compared to MCMC.  However, we avoid the issues with the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal detection function, as will be more familiar to users of classical distance sampling methods. However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \texttt{R-INLA} implementation of INLA \citep{rue_approximate_2009}.  We address this by using a method of iterated model fits based on a first-order Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019}.

We fit the model to point transect \akepa{} data, which differ from line transect methods in that there is a need to account for the increasing area surveyed as distance increases from the transect.  Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process framework is not new and has been applied numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.

@@

In this analysis we assume a half-normal detection function with no covariates on detection parameters to demonstrate the approach in a simplified setting.  The detection function parameters can only be estimated if an assumption is made about $\lambda(\bs)$ otherwise $g(\bs)$ and the intensity are confounded.  The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(\bs)$. The usual approach for two stage models is to assume $\lambda(\bs)$ is constant within each spatial unit, however we are able to relax this assumption slightly to allow $\lambda(\bs)$ to be a linear function within each transect.  Given such an assumption, any observed deviations from uniformity can be attributed to detectability and not to variation in the intensity.

@@

\sloppy The likelihood \eqref{lgcp-likelihood} contains an integral that cannot be solved analytically.  A common approach for fitting log-Gaussian Cox processes in practice is to approximate the likelihood by replacing the integral with a weighted sum.  In this section we describe this approximate likelihood and show how it can be written as a modified Poisson likelihood.  We adapt the approach given in \cite{simpson_going_2016} for a fully observed log-Gaussian Cox process, modified to account for the point transect distance sampling observation process.

To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.  In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes any location within the point transect.  It should be clear from context which is intended by $\bs_k$.   The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$.

@@

The approach taken here is similar to the Berman-Turner device \citep{baddeley_practical_2000, berman_approximating_1992} that is used to fit a wide variety of point process models using GLM software.  The main difference with our approach is that we do not use the assumption that the locations of observed points form a part of the quadrature scheme.

@@

Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.  New linearisation points are found by considering the functional
\begin{equation}
	f(\bar{p}_{\bm{v}}) = \left\{ \argmax_{\bm{u}_i} \: \bar{p}_{\bm{v}}(u_i | \bm{Y}), \quad i = 1, \ldots, n_{\bm{u}} \right\}
\end{equation}
where $n_{\bm{u}}$ is the length of $\bm{u}$.  Each marginal posterior $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ and their modes are estimated during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost for each fit of the linearised model.
We then seek a fixed point of this functional so that $\bm{u}_0 = f(\bar{p}_{\bm{u}_0})$.  This fixed point is identified via an iterative scheme. Given a current choice of linearisation point $\bm{v}$, we estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.  We iterate this procedure until a fixed point is identified within a chosen tolerance.

@@

This section describes the spatial model for the intensity process.  
As mentioned above, the \akepa{} population exhibits spatial heterogeneity for which there is no known explanatory covariate.  Therefore we use a spatially structured random effect to explain this variation.

Using spatially-structured random effects can be computationally expensive.  Often this is due to parameterisations and approximations that lead to dense prior precision matrices which have an  $\mathcal{O}(k^3)$ cost to factorise for an effect with $k$ basis parameters.  A key approach in computationally efficient methods for spatially structured random effects is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).  These result in a sparse precision matrix which can be factorised with a cost of $\mathcal{O}(k^{3/2})$.  We use a GMRF representation based on a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.  This representation is constructed by applying finite element methods to solve the SPDE numerically.

@@

In order to specify priors on the Mat\'ern covariance we use a reparameterisation of $\kappa$ and $\tau$ to a range and variance parameter.  When $\nu = 1$ and the domain is two dimensional then the reparameterisation is $\rho = \sqrt{8} / \kappa$ for the range and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ for the marginal variance (\cite{blangiardo_spatial_2013}, p196).  This parameterisation is used to set penalised complexity priors \citep{simpson_penalising_2017} on $\rho$ and $\sigma^2$ which allow the GRMF effect to shrink towards zero.  For the Mat\'ern field, the base model is the limiting case with $\sigma^2 \rightarrow 0$ and $\rho \rightarrow \infty$, which defines a random field that is almost surely zero everywhere.  The purpose of penalised complexity priors is to place sufficient prior probability on a simple model to allow the random effect to be essentially `turned off' if the model component is not supported by the data.  To specify penalised complexity priors on the SPDE effect we set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.  The value of 130 for the range parameter was selected based on the minimum distance between sampling locations.

@@

The predicted mean of the posterior intensity field is shown in \autoref{fig:intensity-mean-cv}A along with a map of the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid with each prediction cell area approximately 1.7 hectares.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$.  The colour scale is different for each panel.}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data \citep{camp_dsm_2020}.  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity.  There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north.  In the north, estimated intensities are lower, which also contributes to larger CV values and makes the CV map hard to interpret.  The standard deviation map shows stronger overall posterior variability in the south where intensity is larger, most likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.

We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location (Figure \ref{fig:intensity-quantiles}).  As we discuss below, these quantiles are independent summary statistics of the posterior intensity field at each prediction location.  
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_quantiles.png}
		\caption{The predicted posterior intensity quantiles:  A) 0.025 quantile B) 0.975 quantile.  Both plots use the same colour scale.}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}

The posterior detection function (\autoref{fig:det-matern}) shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres and also broadly agrees with a two-stage analysis.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/detfn_and_matern.png}
		\caption{A) Posterior half-normal detection function. B) Posterior Mat\'ern correlation function.  The black line is the mean posterior estimate and the grey shaded area is 95\% credible interval in both plots}
		\label{fig:det-matern}
	\end{center}
\end{figure}
The posterior Mat\'ern correlation function is shown in \autoref{fig:det-matern}; the grey regions show the 95\% credible interval.  We note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).  This is because the correlation function is defined on the log-intensity scale.  On the intensity scale the empirical range of correlation is shorter.

@@

To plot the posterior abundance estimate for the \akepa{} in Hakalau we use a Monte Carlo method based on sampling realisations of the posterior intensity field.  Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$ its corresponding random variable within our Bayesian framework.  Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e. the expected abundance.  To generate a full posterior estimate for abundance we approximate the posterior $\pi(N | \bm{Y})$ by taking $m$ Monte Carlo samples of the posterior intensity field, $\lambda^{(1)}, \ldots, \lambda^{(m)}$ and approximate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)}, \bm{Y})$. Each $\pi(N | \lambda = \lambda^{(i)}, \bm{Y})$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda^{(i)}(\bs)\mathrm{d}\bs$. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[scale=0.525]{figures/N_posterior.png}
		\caption{Posterior abundance.  Grey area marks two Monte-carlo standard errors above and below the mean}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}

@@

The abundance estimates cannot be directly compared to the results in \cite{camp_dsm_2020} since their two-stage model also includes a temporal correlation and we include an area of Hakalau that the authors excluded.  Converting the above abundance estimates to density per hectare for the whole study region allows for some comparison with Figure 2 in \cite{camp_dsm_2020} which suggests that \akepa{} density was roughly between 0.5 and 1.5 birds per hectare in 2002 (considering both the design-based estimate and the spatio-temporal GAM estimate).  In our analysis we find a credible region of roughly 0.9 to 1.7 birds per hectare, which is slightly different but overlaps significantly with the results in \cite{camp_dsm_2020}.
